Food-web structure, stochasticity and response diversity drive temporal stability in complex communities

### 2023-10-19 paper/manuscript.Rmd
Update SEM results

Second, the relation between community stability and biodiversity in variable
environment is contingent on species richness per se, on the effects of species
richness on population stability and asynchrony, and ultimately
on the diversity of species responses to enviromental fluctuations [response
diversity, @ives_stability_1999; @elmqvist_response_2003]. Increasing
species richness has been reported to decrease population stability in
experimental grasslands [@tilman_biodiversity_2006; @rosher_identifying_2011],
while increasing population stability in natural settings
[@houlahan_negative_2018] [add mechanisms?]. In turn, higher species richness
often results in higher asynchrony, through portfolio effects (i.e. statistical
averaging effects due to independant species fluctuations) and compensatory
dynamics (due to species interactions and response
diversity of species due to the environment) [@loreau_species_2008;
@thibaut_understanding_2013]. Systematic increase in asynchrony with species
richness is the mechanism that drives reported positive relationships between
community stability and species richness [ref]. Anthropogenic pressures, such as
eutrophication, has been reported to weakens the relationship between species
richness and asynchony in grasslands [@hautier_eutrophication_2014] but such
empiricial evidence remains scarced. However, stressors such as biotic
homogenization [@buisson_toward_2013; @muthukrishnan_invasive_2020;
@saladin_rapid_2020] and environmental forcing that decrease species response
diversity should dampens the relationships between asynchrony and species
richness [@ives_stability_2000; @thibaut_understanding_2013] due to a decrease
in portefolio effects and compensatory effects. [Talk about the effects of
evenness in temporal variability and evenness in abundance? e.g.,
@zhao_biodiversity_2022; @thibaut_understanding_2013]. Because stressors can
affect both community stability per se and how community stability scales with
species richness, improving our understanding require to study both
[@ives_stability_2007; @hautier_anthropogenic_2015].

Third, the structure of species trophic interactions, that determines the
distribution of biomass fluxes through communities, can also control stability.
including levels of generalism, the distribution of interaction strengths and
the prevalence of omnivory, can also influence stability. The degree of
generalism of predator was shown to modulate stability-richness relationships
[@thebault_trophic_2005]. The strength and distribution of interaction strength
was also shown to be strong driver of community stability, such as stronger
interaction strengths can be destabilizing [@may_will_1972; @mccann_weak_1998]. Omnivory was also
shown to drive higher population stability [@mccann_reevaluating_1997].
The effects of food-web structure on ecosystem stability has been mostly
assessed with small modules [but see @eschenbrenner_diversity_nodate], and we
know very little how food-web structure affects the relationships between
species richness and temporal stability in complex food-webs.

@@

## The model

@@

We added a stochastic natural mortality rate to the species dynamics, that
represent variation in mortality due to variation in environment. Based on the
litterature and previous models [@vasseur_environmental_2007], we assumed that
natural mortality rates scales inversely with the species body mass ($d_i =
d_0M_i^{-1/4}$, ), as all other physiological parameters. We used a basal
natural mortality rate of $d_0 = .4$, as done in previous
studies[@vasseur_environmental_2007].

@@

@@

We generated food-web structure using the niche model [@williams_simple_2000]
with species richness from 10 to 60, and connectance from $0.02$ to $.32$. We
generated food-webs without cannibalistic links and checked for the absence of
disconnected species.

To vary interaction strength and mortality rates in the community,
we varied the Predator-Prey Bodysize Ratio (PPBR) from 1 to 100. We further
varied the functional reponse of consumer from type II to type III (resp. $h=2$
and $h=3$, eq XX) as it was shown to have a tremendous role on stability
[@brose_allometric_2006]. We set $K = XX$ to ensure that consumers were not
limited by the biomass input in the food-web.

@@

We numerically solved the simulation with a $dt$ = .1 during 5000 timesteps
and collected the last 500 timesteps. We set extinction threshold to $10^{-5}$
and if an extinction happened in the last 500 timesteps, we continued to run the
simulation for another 500 timesteps until there was no extinction events.

@@

To further get a mechanistic insight about the effects of stressors and food-web structure on stability,
we partionned asynchrony into statistical averaging effect (SAE) and compensatory
dynamics effect (CPE) [@zhao_biodiversity_2022]. SAE controls for ..., while CPE describes compensatory
dynamics emerging from species interactions and differential responses to
environmental stochasticity.

\begin{equation}
\begin{aligned}
S_{com} = S_{pop} \times SAE \times CPE
(\#eq:scom)
\end{aligned}
\end{equation}

@@

As SAE is further dependant of biomass evenness in food-webs, we
partionned $SAE$ into $SAE_{even}$ and evenness $EVN$, such as $SAE_{even} =
\sqrt{S}$ and $EVN = SAE / SAE_{even}$.

@@

Using a structural equation model, we disantengled the complex effects of
food-web structure, environmental stochasticity and response diversity on the
different facets of temporal stability of biomass (i.e. population stability,
statistical averaging effects and compensatory dynamic effects). We set the
paths of the path of the model according to theorical expectations. We tested
the effects of species richness, connectance, and the Predator-Prey size ratio
on secondary structure of the food-webs, the average trophic level, the average
omnivory and the interaction strength. We also tested the effects of  on  We
expected that richness affect population stability by modifying the average
population biomass for exemple [@thibaut_understanding_2013]. Grounded in
theoretical ecological, we expected that average trophic level, the average
omnivory and the interaction strength modulated population stability
[@mccann_weak_1998; @mccann_reevaluating_1997; @shanafelt_stability_2018].
Previous studies have also the Predator-Prey size ratio modulate population
stability [@brose_allometric_2006]. Finally, the effect of enviromental
stochasticity was assumed to be mainly associated to population stability (i.e.
by increasing amplitude of population fluctuations).

Regarding asynchrony, we tested the effects of species richness, connectance,
weighted average trophic level and interaction strength on the evenness of
biomass which in turn affect total statistical averaging effects. We expected
that those same food-web structural properties also affected compensatory
dynamics due to interactions ($CPE_{int}$). Finally, we tested the effects of
species richness and species cross-correlation in mortality rates on the
compensatory dynamics due to stochastic fluctuations ($CPE_{env}$). We completed
the structural equation model with the mathematical relationship linking
compensatory dynamics and statistical averaging effects to asynchrony, and from
asynchrony and population stability to community stability.

@@

We found that food-web structure had significant effect on temporal population
stability such as average trophic level and type III functional response (i.e. hill exponent = 3)
increase population stability (resp. $r_\delta$= `r rpt` and `r rph`, $r_\delta$
being the standardized coefficient, Fig. \@ref(fig:sem)). However the strength
of environmental stochasticity had a strong negative effect on population
stability ($r_\delta$= `r rpe`).

We further found that food-web structure had a strong impact on statistical
averaging effects, which raises from independant species fluctuations weigthed
the evenness of their biomass. As expected the mathematical relationship (eq.
XX), species richness was a main positive driver of statistical averaging effect
($r_\delta$= `r rsr`, Fig. \@ref(fig:sem)), thereby increasing asynchrony.
Average trophic level and average interaction strength had strong negative
effects on statistical averaging effects (resp. $r_\delta$= `r rst` and
`r rsi`), thereby downplaying the portefolio effect raised from species
richness. In turn, we found that average interaction strength had a small
positive effect on compensatory dynamics ($r_\delta$= `r rci`, Fig.
\@ref(fig:sem)). Cross-correlation among species mortality rates had a
strong negative effect on compensatory dynamics ($r_\delta$= `r rco`), meaning
that the loss of response diversity result in a strong decrease in compensatory
dynamics.

@@

rscp <- get_eff(y = "stab_com", x = "pop_stab", tab = sem_eff_tab)
rsca <- get_eff(y = "stab_com", x = "async", tab = sem_eff_tab)

rac <- get_eff(y = "async", x = "cpe")
ras <- get_eff(y = "async", x = "sae_total")

@@

Deriving the sum of direct and indirect effects from the structural equation
model, we show that both enviromental stochasticity and cross-correlation among
species had by far the most negative effect on community stability (XX, Fig.
\@ref(fig:semtot)). Average trophic level, type III functional response and
predator prey bodysize ratio had total positive effects on community stability.
Those positive effects are raising from sometimes contrasting effects on
population stability and asynchrony such as Average trophic level, type III
functional response and predator prey bodysize ratio which have positive effects
on population stability (resp. XXX) but a negative one on asynchrony (resp. XXX).

@@

Our results shows that community stability in complex food-webs is mainly driven
by population stability rather than synchrony, confirming previous results in
empirical data [@danet_species_2021]. We further found that average trophic
level and functional response of type III increased average population
stability, confirming previous results [@brose_allometric_2006;
@shanafelt_stability_2018; @danet_species_2021;
@eschenbrenner_diversity_nodate]. Higher average trophic level stabilises
populations because higher predator are themselves more stable and because they
control the amplitude of the fluctuations of their prey [REF, we can test
that?]. Furthermore, average omnivory was strongly correlated with average
trophic level, suggesting that average omnivory also stabilise population
stability, in line with previous theoretical results in small modules
[@mccann_reevaluating_1997]. In turn, functional response of type III decreases
the feeding rates of consumers when their preys have a low biomass which leads
to a less dramatic fluctuations of prey biomass [@uszco_effects_2017;
@brose_allometric_2006]. We however found no big effect of species richness on
population stability, which has been reported to be negative, positive or
neutral in previous studies [@ruijven_contrasting_2007;
@houlahan_negative_2018].

>>>>

Second, the relation between community stability and biodiversity in variable
environment is contingent on species richness per se, on the effects of species
richness on population stability and asynchrony, and ultimately
on the diversity of species responses to enviromental fluctuations [response
diversity, @ives_stability_2000; @elmqvist_response_2003]. Increasing
species richness has been reported to decrease population stability in
experimental grasslands [@tilman_biodiversity_2006; @roscher_identifying_2011],
while increasing population stability in natural settings
[@houlahan_negative_2018] [add mechanisms?]. In turn, higher species richness
often results in higher asynchrony, through portfolio effects (i.e. statistical
averaging effects due to independant species fluctuations) and compensatory
dynamics (due to species interactions and response
diversity of species due to the environment) [@loreau_species_2008;
@thibaut_understanding_2013]. Systematic increase in asynchrony with species
richness is the mechanism that drives reported positive relationships between
community stability and species richness [ref]. Anthropogenic pressures, such as
eutrophication, has been reported to weakens the relationship between species
richness and asynchony in grasslands [@hautier_eutrophication_2014] but such
empiricial evidence remains scarced. However, stressors such as biotic
homogenization [@buisson_toward_2013; @muthukrishnan_invasive_2020;
@saladin_rapid_2020] and environmental forcing that decrease species response
diversity should dampens the relationships between asynchrony and species
richness [@ives_stability_2000; @thibaut_understanding_2013] due to a decrease
in portefolio effects and compensatory effects. [Talk about the effects of
evenness in temporal variability and evenness in abundance? e.g.,
@zhao_biodiversity_2022; @thibaut_understanding_2013]. Because stressors can
affect both community stability per se and how community stability scales with
species richness, improving our understanding require to study both
[@ives_stability_2007; @hautier_anthropogenic_2015].

Third, the structure of species trophic interactions, that determines the
distribution of biomass fluxes through communities, can also control stability.
including levels of generalism, the distribution of interaction strengths and
the prevalence of omnivory, can also influence stability. The degree of
generalism of predator was shown to modulate stability-richness relationships
[@thebault_trophic_2005], and generalist predators can synchronise the dynamics
of their preys[@raimondo_interspecific_2004]. The strength and distribution of
interaction strength was also shown to be strong driver of community stability,
such as stronger interaction strengths can be destabilizing [@may_will_1972;
@mccann_weak_1998]. Omnivory was also shown to drive higher population stability
[@mccann_reevaluating_1997]. The effects of food-web structure on ecosystem
stability has been mostly assessed with small modules [but see
@eschenbrenner_diversity_nodate], and we know very little how food-web structure
affects the relationships between species richness and temporal stability in
complex food-webs. So understanding the effects of stressors on community
stability and on community stability-richness relationships necessitate to
account for food-web structure.

@@

## Bioenergenetic model

@@

@@

We added a stochastic natural mortality rate to the species dynamics,
representing a fluctuating environment. Based on the litterature and previous
models [@vasseur_environmental_2007], we assumed that
natural mortality rates scales inversely with the species body mass ($d_i =
d_0M_i^{-1/4}$, ), as all other physiological parameters. We used a basal
natural mortality rate of $d_0 = .4$, as done in previous
studies[@vasseur_environmental_2007].

@@

We generated food-web structure using the niche model [@williams_simple_2000]
with an initial species richness from $10$ to $60$, and connectance from $0.02$
to $.32$. We generated food-webs without cannibalistic links and discarded
food-webs that contained disconnected species.

To vary interaction strength and mortality rates in the community,
we varied the Predator-Prey Bodysize Ratio (PPBR) from 1 to 100. We further
varied the functional reponse of consumer from type II to type III (resp. $h=2$
and $h=3$, eq XX) as it was shown to have a tremendous role on stability
[@brose_allometric_2006]. We set $K = XX$ to ensure that consumers were not
limited by the biomass input from primary producers.

@@

We numerically solved the simulations with a $dt$ = .1 during 5000 timesteps
and collected the last 500 timesteps. We set extinction threshold to $10^{-5}$
and if an extinction happened in the last 500 timesteps, we continued to run the
simulation for another 500 timesteps until there was no extinction events.

@@

To further get a mechanistic insight about the effects of stressors and food-web
structure on stability, we partionned asynchrony into statistical averaging
effect (SAE) and compensatory dynamics effect (CPE) [@zhao_biodiversity_2022].
SAE quantities the portfolio effect emerging from independant species
fluctuations, while CPE describes compensatory dynamics emerging from species
interactions and differential responses to environmental stochasticity.

\begin{equation}
\begin{aligned}
S_{com} = S_{pop} \times SAE \times CPE
(\#eq:scom2)
\end{aligned}
\end{equation}

@@

As SAE is further dependant on the evenness of species variance, we
partionned $SAE$ into $SAE_{even}$ and evenness $EVN$, such as $SAE_{even} =
\sqrt{S}$ and $EVN = SAE / SAE_{even}$.

@@

Using a structural equation model, we disantengled the complex effects of
food-web structure, environmental stochasticity and response diversity on the
different facets of temporal stability of biomass (i.e. population stability,
statistical averaging effects and compensatory dynamic effects). We set the
paths of the path of the model according to theorical expectations. We tested
the effects of initial species richness, initial connectance, Predator-Prey size
ratio on secondary structure of the food-webs, i.e. realized species richness
and connectance, the average trophic level, the average omnivory and the
interaction strength. Further, we expected those food-web structure to alter
population stability and synchrony. Grounded in theoretical ecology, we expected
that higher average trophic level, average omnivory, lower interaction strength,
and type III functional response (relative to type II) increase population
stability [@mccann_weak_1998; @mccann_reevaluating_1997;
@shanafelt_stability_2018]. Higher Predator-Prey size ratio was also expected to
increase population stability [@brose_allometric_2006]. Finally, the effect of
enviromental stochasticity was assumed to be mainly associated to population
stability (i.e. by increasing amplitude of population fluctuations).

Regarding asynchrony, we tested the effects of species richness, connectance,
average trophic level and interaction strength on statistical averaging effects.
We expected that those same food-web structural properties affected
compensatory dynamics due to interactions ($CPE_{int}$). Finally, we tested the
effects of species richness and species cross-correlation in mortality rates on
the compensatory dynamics due to stochastic fluctuations ($CPE_{env}$). We
completed the structural equation model with the mathematical relationship
linking compensatory dynamics and statistical averaging effects to asynchrony,
and from asynchrony and population stability to community stability.

@@

We found that food-web structure had significant effect on temporal population
stability such as average trophic level and type III functional response (i.e.
hill exponent = 3) increase population stability (resp. $r_\delta$= `r rpt` and
`r rph`, $r_\delta$ being the standardized coefficient, Fig. \@ref(fig:sem)).
In turn, the variance of environmental stochasticity had a strong negative effect
on population stability ($r_\delta$= `r rpe`).

We further found that food-web structure had a strong impact on statistical
averaging effects, which raises from independant species fluctuations weigthed
the evenness of their variance. As expected the mathematical relationship (eq.
XX), species richness was a main positive driver of statistical averaging effect
($r_\delta$= `r rsr`, Fig. \@ref(fig:sem)), thereby increasing asynchrony.
Average trophic level and average interaction strength had strong negative
effects on statistical averaging effects (resp. $r_\delta$= `r rst` and
`r rsi`), thereby downplaying the portefolio effect raised from species
richness. In turn, we found that average interaction strength had a small
positive effect on compensatory dynamics ($r_\delta$= `r rci`, Fig.
\@ref(fig:sem)). Cross-correlation among species mortality rates (i.e. lower
response diversity) had a strong negative effect on compensatory dynamics
($r_\delta$= `r rco`), meaning that the loss of response diversity result in a
strong decrease in compensatory dynamics.

@@

rscp <- get_eff(y = "stab_com", x = "pop_stab")
rsca <- get_eff(y = "stab_com", x = "async")
#rscp <- get_eff(y = "stab_com", x = "pop_stab", tab = sem_eff_tab)
#rsca <- get_eff(y = "stab_com", x = "async", tab = sem_eff_tab)

rac <- get_eff(y = "async", x = "cpe")
ras <- get_eff(y = "async", x = "sae_total")
#rac <- get_eff(y = "async", x = "cpe")
#ras <- get_eff(y = "async", x = "sae_total")

@@

```{r}
rtsrho <- get_eff(y = "stab_com", x = "rho", effect_type = "total")
rtsenv <- get_eff(y = "stab_com", x = "env_stoch", effect_type = "total")

rtsr <- get_eff(y = "stab_com", x = "richness", effect_type = "total")
rtst <- get_eff(y = "stab_com", x = "w_avg_tlvl", effect_type = "total")
rtsz <- get_eff(y = "stab_com", x = "Z", effect_type = "total")
rtsh <- get_eff(y = "stab_com", x = "h", effect_type = "total")
rtsc <- get_eff(y = "stab_com", x = "ct_alive", effect_type = "total")
rtsi <- get_eff(y = "stab_com", x = "avg_int_strength", effect_type = "total")

rtpr <- get_eff(y = "pop_stab", x = "richness", effect_type = "total")
rtpt <- get_eff(y = "pop_stab", x = "w_avg_tlvl", effect_type = "total")
rtpz <- get_eff(y = "pop_stab", x = "Z", effect_type = "total")
rtph <- get_eff(y = "pop_stab", x = "h", effect_type = "total")
rtpc <- get_eff(y = "pop_stab", x = "ct_alive", effect_type = "total")
rtpi <- get_eff(y = "pop_stab", x = "avg_int_strength", effect_type = "total")

rtar <- get_eff(y = "async", x = "richness", effect_type = "total")
rtat <- get_eff(y = "async", x = "w_avg_tlvl", effect_type = "total")
rtaz <- get_eff(y = "async", x = "Z", effect_type = "total")
rtah <- get_eff(y = "async", x = "h", effect_type = "total")
rtac <- get_eff(y = "async", x = "ct_alive", effect_type = "total")
rtai <- get_eff(y = "async", x = "avg_int_strength", effect_type = "total")
```

Deriving the sum of direct and indirect effects from the structural equation
model, we show that both enviromental stochasticity and cross-correlation among
species had by far the most negative effect on community stability (resp.
$r_\delta$= `r rtsenv` and `r rtsrho`, Fig. \@ref(fig:semtot)). Species richness
had a low overall positive total effect on community stability (`r rtsr`)
because its strong total positive effect on asynchony (`r rtar`) was downplayed
by its total negative effect on population stability (`r rtpr`) and that the
weigtht of population stability on community stability is twice higher than the
one of asynchony. Average trophic level, type III functional response and
predator prey bodysize ratio had positive effects on community stability in
total (resp. `r rtst`, `r rtsh`, `r rtsz`, Fig. \@ref(fig:semtot)), driven by
their positive effects on population stability that were weakly counterbalanced
by their negative effects on asynchrony. In turn, connectance and average
interaction strength had low total negative effects on community stability (resp.
`r rtsc` and `r rtsi`, Fig. \@ref(fig:semtot)). While connectance had negative
effects both on population stability and asynchrony, average interaction
strength had a low total positive effect on population stability counterbalanced
by a stronger negative effect on asynchrony.

@@

Our results shows that community stability in complex food-webs is mainly driven
by population stability rather than synchrony, confirming previous results in
empirical data [@danet_species_2021]. We further found that average trophic
level and functional response of type III increased average population
stability, confirming previous results [@brose_allometric_2006;
@shanafelt_stability_2018; @danet_species_2021;
@eschenbrenner_diversity_nodate]. Higher average trophic level stabilises
populations because higher predator are themselves more stable and because they
control the amplitude of the fluctuations of their prey [REF, we can test
that?]. Furthermore, average omnivory was strongly correlated with average
trophic level, suggesting that average omnivory also stabilise population
stability, in line with previous theoretical results in small modules
[@mccann_reevaluating_1997]. In turn, functional response of type III decreases
the feeding rates of consumers when their preys have a low biomass which leads
to a less dramatic fluctuations of prey biomass [@uszko_effects_2017;
@brose_allometric_2006]. We however found no big effect of species richness on
population stability, which has been reported to be negative, positive or
neutral in previous studies [@ruijven_contrasting_2007;
@houlahan_negative_2018].
